---
title: "Overrepresentation in Jails in U.S. Counties"
author: "Kensho Gendzwill"
date: '2022-11-21'
output: html_document
---

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
source("../source/a4-helpers.R")
source("../source/analysis.R")
```

## Abstract 
In this report, I demonstrate that black people disproportionately make up more
of the jail population relative to the general county population in a vast
majority of counties. I also show that black and latinx people routinely make up
majority jail populations in counties where they are minorities, while the same
is rarely true for white people.

## Introduction 
In this project, my analysis focused on how certain races were misrepresented in
each U.S. county's jail system. In this report, I have demonstrated that black
and latinx people tend to be overrepresented in jails when compared to the 
county population. Black and latinx people have a long history of being targeted
and oppressed by the U.S. legal system, and this report provides analysis in
support of this claim. The primary stakeholders of this report are policymakers
and law enforcement officers, especially those who represent the counties that
this report has found inequalities in. I hope they can use the results of this
report to revise law and enforcement so that race can become less of a
determinant for who is jailed and who is not. The secondary stakeholders,
by extension, is the black and latinx communities that are affected by
disproportionate jailing.
The questions I answered were:  
-How has the total U.S. jail population increased over the years?  
-How has the total jail population in certain counties increased over the years?  
-How do the percentages of the black population in counties and in their jails
compare?  
-Which counties have a different majority demographic between the county
population and the jail population?

## Data Summary 
In my analysis, I chose to analyze which counties misrepresented certain races
in their prisons. My first attempt at this was to compare the percentage of
black people in each county and their jails. My results showed me that in
`r black_pop_county_count` counties, more black people represented the jail
population than they did in the county population. This means that black people
were overrepresented in jails in `r black_pop_diff_pct`% of all U.S. counties.  
Secondly, I compared the majority demographic of each U.S. county with the
majority demographic in each county's jail population. This analysis showed that
`r most_pop_combined_count` out of all `r county_count_2018` counties had a majority
jail demographic that was different from its majority county demographic. This
amounts to `r most_pop_diff_pct`% of all counties.

## Growth of the U.S. Prison Population
```{r, fig.width=10, fig.height=4, echo=FALSE, fig.cap="Increase of Jail Population in U.S. - a bar chart showing the trend of the total incarcerated population in the U.S. from 1970 to 2018."}
plot_jail_pop_for_us()
```
The graph shows a trend of mostly increasing incarcerations since 1970. The
incarcerated population was at an all time low around 1971, and steady grew
until it reached its peak in 2009. Since then, the total incarceration
population has shrunk by around 50,000 people. 

## Growth of the Prison Population by State
```{r, fig.width=10, fig.height=4, echo=FALSE, fig.cap="Increase of Jail Population in U.S. States - a line chart showing the trend of the total incarcerated population in U.S. states from 1970 to 2018. States to be displayed are specified in function call.", message=FALSE}
plot_jail_pop_by_state(c('AZ', 'CO', 'ID', 'MT', 'NV', 'NM', 'UT', 'WY'))
```
The graph shows a trend of incarcerations in the mountain states since 1970. The
states were chosen so the incarceration trends could be analyzed across a single
U.S. region. According to the graph, the incarcerated population was at an all
time low around 1971 for most states, and steady grew until it reached its peak
in 2009. Since then, most states in the region have seen a decrease in total
incarcerated population.

## Comparing the Percentages of the Black Population in County and Jail
```{r, fig.width=8, fig.height=8, echo=FALSE, fig.cap="Percentage of the Black Population in County and Jail - a scatter plot showing the percentage of the black population in each county and its jails (2018)."}
plot_black_pop()
```
The graph shows a scatter plot of the percentage of the black population in each
county and its jails from 2018. Each dot represents a county in the U.S. The
dots are color-coded by region. The line y=x is plotted for reference. If all of
the points were along this line, we would be able to say that there is no
correlation between being black and being in jail. However, a majority of the
points are above the line, meaning that black people are overrepresented in
jails. The problem especially seems to be bad in midwestern states, where low
county populations of black people are juxtaposed by high jail populations of
black people. 

## Comparing Majority Demographics in U.S. Counties and their Jails
```{r, fig.width=10, fig.height=5, echo=FALSE, fig.cap="Majority Demographic in U.S. Counties and their Jails - the first plot shows a map of the U.S., each county color-coded with its majority demographic. The second plot shows the same map, color-coded with each county's majority jail demographic"}
plot_most_pop()
plot_most_pop_jail()
```
The two maps show where the majority demographic of a county is misrepresented
in its jails. Namely, large swaths of counties in the southeast seem to
disproportionately jail black people. The same holds true for latinx people in
the southwest. In general, it seems rare for white people to represent a majority
in jails where they are not the majority in the county population, while it is
more common for black and latinx people to represent a jailed majority in
counties where they are a minority.
